### 2025-09-19 paper/abstract.tex

\begin{abstract}
We introduce the idea of \emph{transparent documents}, web-based data-driven scholarly articles which allow
readers to explore the relationship to the underlying data by hovering over fragments of text. We present an
agent-based LLM framework for authoring transparent documents, building on recent developments in data
provenance for general-purpose programming languages. Our implementation uses Fluid, an open source functional
programming language with a provenance-tracking runtime, as a target platform. Our tool consists of two LLM
agents which support a human author during the creation of a transparent document. A \SuggestionAgent helps
identify fragments of text which could plausibly be computed from data, including numerical values selected
from records or computed by aggregations like sum and mean, comparatives and superlatives like ``better than''
and ``largest'', trend-adjectives like ``growing'', and other idiomatic quantitative or semi-quantitative
phrases. An \InterpretationAgent, given such a fragment, then attempts to synthesise a suitable Fluid query
over the data which will generate the target string. The resulting expression is spliced into the source code
for an interactive web page, turning the static text fragment into an interactable data-driven element able to
reveal the data that underwrites the natural language claim. We evaluate our approach on a subset of SciGen,
an open source dataset consisting of tables from scientific articles and their corresponding descriptions,
which we extend with hand-generated counterfactual test cases to evaluate how well machine-generated
expressions generalise in the presence of changes to the data. Our results show that some state-of-the-art
models are often able to synthesise compound expressions that are extensionally compatible with our gold
solutions.
\end{abstract}

>>>>

\begin{abstract}
We introduce \emph{transparent documents}, interactive web-based scholarly articles which allow readers to
explore the relationship to the underlying data by hovering over fragments of text. We present an agent-based
LLM framework for authoring transparent documents, building on recent developments in data provenance for
general-purpose programming languages. Our implementation uses Fluid, an open source functional programming
language with a provenance-tracking runtime, as a target platform. Our tool consists of two LLM agents which
support a human author during the creation of a transparent document. A \SuggestionAgent helps identify
fragments of text which could plausibly be computed from data, including numerical values selected from
records or computed by aggregations like sum and mean, comparatives and superlatives like ``better than'' and
``largest'', trend-adjectives like ``growing'', and other idiomatic quantitative or semi-quantitative phrases.
An \InterpretationAgent, given such a fragment, then attempts to synthesise a suitable Fluid query over the
data which will generate the target string. The resulting expression is spliced into the source code for an
interactive web page, turning the static text fragment into an interactable data-driven element able to reveal
the data that underwrites the natural language claim. We evaluate our approach on a subset of SciGen, an open
source dataset consisting of tables from scientific articles and their corresponding descriptions, which we
extend with hand-generated counterfactual test cases to evaluate how well machine-generated expressions
generalise in the presence of changes to the data. Our results show that some state-of-the-art models are
often able to synthesise compound expressions extensionally compatible with our gold solutions.
\end{abstract}

### 2025-09-19 paper/sec/introduction.tex

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making them hard for reviewers to
check directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but current LLM
interfaces do not support direct interrogation of visual or other outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,bond25}, on the other hand,
takes a more infrastructural approach to the transparency problem. These approaches link computed outputs to
their data sources directly, by tracking dependency information. This allows visual outputs to support
\emph{provenance queries}, user interactions (e.g. mousing over a visual element) that reveal how visual
features relate to data. The advantage of this approach is that the interactions are provided automatically
via trusted infrastructure, namely a query language or general-purpose programming language which tracks how
data flows through a computation. However, these approaches are limited to outputs computed from data, such as
visualisations. What is missing is a way to extend these ``direct interrogation'' features to natural language
itself, where the main claims of most scholarly articles are actually made.

In this paper, we address this gap by combining these two approaches, leveraging the ability of LLMs to
understand technical language and synthesise queries over data, plus the provenance-tracking infrastructure of
an open source programming language called Fluid (\url{https://f.luid.org/}) designed originally for
transparent visualisations~\citep{perera22,bond25}. We show how these two technologies can together support
the creation and deployment of \emph{transparent documents}, web-based scholarly articles where key
quantitative claims expressed in natural language --- such as the claim that one software system is faster
than another at solving a problem --- are computed from the relevant data, and that support interactive
provenance queries that reveal to a reader or reviewer the specific data elements that support the claim.

\figref{scigen-example-website}, adapted from \cite{zhang18}, shows how a transparent document appears to a
user. \rpnote{Make important point about counterfactuals/generalisability}

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base is open and accessible.
A key challenge lies in tracing specific claims back to the underlying data. In peer review, for example,
empirical claims typically lack author-supplied links to data, making them hard for reviewers to check
directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to simple
errors in data management or analysis~\citep{hu25}. The use of large language models (LLMs) to interpret
scholarly documents has seen considerable attention recently, from fact-checking~\citep{abu-ahmad25} to
interpretation of charts and figures~\citep{roberts24}, but current LLM interfaces do not support direct
interrogation of visual or other outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,bond25}, on the other hand,
takes a more infrastructural approach to tracing outputs back to data. These approaches link computed outputs
to their data sources directly, by tracking dependency information. This allows visual outputs to support
\emph{provenance queries}, user interactions (e.g. mousing over visual elements) that reveal how output
features relate to data. The advantage of this approach is that the relationships to data sources are exposed
automatically via trusted infrastructure, typically a query language or general-purpose programming language
which tracks how data flows through a computation. However, these approaches are limited to outputs computed
from data, such as visualisations. What is missing is a way to extend these ``direct interrogation'' features
to natural language itself, where the main claims of most scholarly articles are actually made.

In this paper, we address this gap by combining two complementary approaches: the ability of LLMs to
understand technical language and synthesise queries over data, plus the provenance-tracking infrastructure of
an open source programming language called Fluid (\url{https://f.luid.org/})~\citep{perera22,bond25}.
Together, these two technologies enable the creation of \emph{transparent documents}, web-based scholarly
articles with two key transparency features:
\begin{enumerate}
\item \textbf{Data-driven:} Quantitative statements expressed in natural language --- e.g.~that system $X$ is
faster than system $Y$ on some task --- are computed from the relevant data, rather than occurring
merely as static fragments of text.

\item \textbf{Data linking:} Readers and reviewers can interactively trace such claims back to the specific
data elements that support them, through embedded provenance queries.
\end{enumerate}

\figref{scigen-example-website}, generated from our implementation, illustrates these two features. The upper
section shows an excerpt from a research paper comparing text encoding techniques~\citep{zhang18}. When a
reader hovers over the phrase ``does not further improve'', the corresponding data entries are highlighted in
blue on the left. Other fragments (e.g. ``better than'', ``further improvements'') that refer to the same data
are also marked, allowing readers to explore supporting and contrasting evidence. The lower section shows a
counterfactual situation where we imagine the authors' experiments had produced different results: here the
phrase ``does not further improve'' is replaced by the ``further improves''. Transparent documents thus turn
static text into interactable, data-driven content able to expose the evidential basis of scholarly claims.
